---
title: "NYC Shooting Incidents"
author: "Nicholas Anadon"
date: "2025-04-12"
output: html_document
---

-Data set is NYPD Shooting Incident Data(Historic) last updated on 4/26/2024 on catalog.data.gov.

-Data entries range from 1/1/2006 to 12/29/2023

-What boroughs have more incidents than others and how many total incidents are there across all boroughs?



```{r setup, include = FALSE}
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
```




```{r}
#This code starts the cleaning process
nyshooting <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
nyshooting <- read.csv(nyshooting)

#Changing weird values to UNKNOWN so most data is not lost
sore_val <- c("", "(null)", "1028", "1020", "940", "224", "1022")

#Changing names of columns to easier names
nyshooting <- rename(nyshooting, ID_Key = "INCIDENT_KEY", Date = "OCCUR_DATE", Time = "OCCUR_TIME", Boro = "BORO", Inci_Loc = "LOCATION_DESC", Death = "STATISTICAL_MURDER_FLAG",
                     Perp_Age = "PERP_AGE_GROUP", Perp_Sex = "PERP_SEX", Perp_Race = "PERP_RACE",
                     Vic_Age = "VIC_AGE_GROUP", Vic_Sex = "VIC_SEX", Vic_Race = "VIC_RACE", Precinct = 
                       "PRECINCT", Jur_Code = "JURISDICTION_CODE") %>%
 select(-"LOC_OF_OCCUR_DESC", -"LOC_CLASSFCTN_DESC", -"X_COORD_CD", -"Y_COORD_CD",
        -"Latitude", -"Longitude", -"Lon_Lat", -"ID_Key", -"Inci_Loc") %>%
  mutate(across(where(is.character),
                ~ifelse(.x %in% sore_val, "UNKNOWN", .x)))

nyshooting <- na.omit(nyshooting)
nyshooting$Date <- as.Date(nyshooting$Date, format = "%m/%d/%Y")
nyshooting$Death <- as.logical(nyshooting$Death)
nyshooting$Jur_Code <- as.factor(nyshooting$Jur_Code)
nyshooting$Boro <- as.factor(nyshooting$Boro)
nyshooting$Precinct <- as.factor(nyshooting$Precinct)
nyshooting$Perp_Age <- as.factor(nyshooting$Perp_Age)
nyshooting$Perp_Sex <- as.factor(nyshooting$Perp_Sex)
nyshooting$Perp_Race <- as.factor(nyshooting$Perp_Race)
nyshooting$Vic_Age <- as.factor(nyshooting$Vic_Age)
nyshooting$Vic_Sex <- as.factor(nyshooting$Vic_Sex)
nyshooting$Vic_Race <- as.factor(nyshooting$Vic_Race)
#Variable Changes
#Date: character -> date
#Death: character -> logical
#All other variables: character -> factor

#Only 2 entries out of 28,562 entries contained NA values. Values are omitted
#Date allows for quantitative analysis over a selected time
#Logical for death illustrates a binary choice between survival and death of victim
#Factor variables allow for easier categorical variable analysis

head(nyshooting)
```




```{r}
#Transformed data set nydeath of how many incidents occurred on each day
nydeath <- nyshooting %>%
  arrange(Date)

nydeath <- nydeath %>%
  group_by(Date, Death) %>%
  summarise(count = n(), .groups = "drop")

head(nydeath)

#Condensed bar chart where survival incidents stack on death incidents
#Total incidents for a given day are illustrated by the top point of survival bars
nydeath %>%
  ggplot(aes(x = Date, y = count, fill = Death)) +
  geom_col(position = "stack") +
  labs(title = "Daily Shooting Incidents in New York City",
       x = "Date", y = "Number of Incidents") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 year") +
  scale_y_continuous(limits = c(0, 50)) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "darkgreen"),
                    name = "Outcome",
                    labels = c("FALSE" = "Survival", "TRUE" = "Death")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))
```



```{r}
#Transformed cumulative data set nydeath_cuml that shows the total amount of incidents over time
nydeath_cuml <- nydeath %>%
  arrange(Date) %>%
  group_by(Death) %>%
  mutate(cumulative_count = cumsum(count))

#Condensed area chart of total amount of incidents over time
#Same stacked bar rule as the stacked bar chart
#Blue line: Linear line of best fit for survival incidents
#Cyan line: Linear line of best fit for death incidents
nydeath_cuml %>%
  ggplot(aes(x = Date, y = cumulative_count, fill = Death, color = Death)) +
  geom_area(position = "stack", color = NA) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 2) +
  labs(title = "Daily Shooting Incidents in New York City",
       x = "Date", y = "Number of Incidents",
       fill = "Outcome", color = "Outcome") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 year") +
  scale_y_continuous(limits = c(0, 30000)) +
  scale_color_manual(values = c("TRUE" = "cyan", "FALSE" = "blue"),
                     name = "Outcome",
                     labels = c("TRUE" = "Death", "FALSE" = "Survival")
                     ) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "darkgreen"),
                    name = "Outcome",
                    labels = c("TRUE" = "Death", "FALSE" = "Survival")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
#Transformed data set nyboro that shows the amount of incidents within each NYC borough
nyboro <- nyshooting %>%
  arrange(Boro)

nyboro <- nyboro %>%
  group_by(Boro, Death) %>%
  summarise(count = n(), .groups = "drop")

nyboro

#Bar chart showing incidents in boroughs across the data's timeframe
nyboro %>%
  ggplot(aes(x = Boro, y = count, fill = Death)) +
  geom_col(position = "dodge", color = 'black') +
  scale_y_continuous(limits = c(0, 10000)) +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "darkgreen"),
                               name = "Outcome",
                               labels = c("FALSE" = "Survival", "TRUE" = "Death")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(title = "Shooting Incidents in NYC Boroughs 1/1/2006-12/29/2023",
       x = "Borough",
       y = "Number of Incidents",
       fill = "Death")
```


```{r}
#Linear model of deaths as time passes
mod <- lm(count ~ Date, data = nydeath)
mod
summary(mod)
```
**Data Findings:**
Brooklyn was the borough with the largest number of incidents, but that does not 
necessarily mean that it is the most dangerous borough to live in. There were a
couple days where the number of incidents vastly outnumbered a calculated average,
but for the most part the graph is consistent with incidents. The total number of
incidents recorded from the clean data total to 23,034 incidents from January 1st, 2006 
to December 29th, 2023.

**Data Bias:**
There are differing law enforcement standards within each NYC borough, so the results
may not show the true danger within each area. Death is a logical argument in the cleaned data set, so
we do not know the extent of how much victims were injured before they died or survived.
Race and age of attackers and victims were not variables of interest for what I wanted to
uncover. Bias statements to consider when looking over the cleaned data and analysis.
```{r}
write.csv(nyshooting, "nyshooting_cleaned.csv")
```

```{r}
sessionInfo()
```