---
title: "US Covid-19 Cases and Deaths(1/23/20-3/9/23)"
author: "Nicholas Anadon"
date: "2025-04-26"
output: html_document
---

```{r}
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
```

***Instructor Credit and Data Introduction:***
Dr. Jane Wall was a sensational instructor for "Data Science as a Field" and
per her permission on the final project, she has granted permission
for the clean-up code of "COVID-19 Data Repository by the Center for Systems 
Science and Engineering (CSSE) at Johns Hopkins University" on Github.com. This
project gives a cleaned version of COVID-19 cases and deaths strictly within
the United States. The time period of these cases and deaths reported starts at
January 23, 2020 to March 9, 2023. The goal is to analyze what months had more
cases and deaths than others and what are the summary statistics behind the 
months with the most daily differences in COVID cases and deaths.

***Data set URL:***
https://github.com/pomkos/john-hopkins-covid-data

```{r}
US_cases <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv"
US_deaths <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv"

US_cases <- read.csv(US_cases)
US_deaths <- read.csv(US_deaths)
```

## Dr. Wall's Cleanup Code and Summary of Code:

```{r}
#Values of US cases
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  select(-c(Lat, Long_))
US_cases$date <- as.Date(sub("^X", "", US_cases$date), format = "%m.%d.%y")


#Values of US deaths
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  select(-c(Lat, Long_))
US_deaths$date <- as.Date(sub("^X", "", US_deaths$date), format = "%m.%d.%y")

#Both US cases and US deaths have the same variables: 
#Admin2 - City within the State
#Province_State - US State
#Country_Region - Always US(Important to note that this data set only has US
#data while global statistics were also recorded)
#Combined_Key - City + State
#Population - Total population recorded at the start date
#date - Date


#full_join() combines the variables not included in both data sets.

#cases - Cumulative sum of cases within a state as time progresses
#deaths - Cumulative sum of deaths within a state as time progresses
US <- US_cases %>%
  full_join(US_deaths)

#Eliminating negative values
US <- US %>%
  mutate(
    cases = pmax(cases, 0),
    deaths = pmax(deaths, 0)
  )

tail(US)
summary(US)
```


```{r}
#Combining data from all cities within states
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

tail(US_by_state)

#Combining data from states within the United States
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

#deaths_per_mill - Number of deaths expected out of 1 million people in relation to the population
tail(US_totals)
summary(US_totals)
```

```{r}
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
#It's important to note that some values in new_cases and new_deaths are 
#negative. This is okay, because we are looking at the difference between cases
#of one day and the day prior regardless if someone already had the disease or
#had just caught the disease that day. Thinking in real time, if there were 
#fewer cases today than yesterday, the result would be negative.
```

## Nicholas Anadon's Data Fixes and Summaries:

```{r}
#Only one data value contained an NA value so it has been omitted

#Months and Seasons are primary variables of interest

#For simplicity in the next few data visualizations the seasons cut off after 
#the month in which the next season begins
US_totals <- na.omit(US_totals)
US_totals$month <- month(US_totals$date, label = TRUE)
US_totals$month_name <- month(US_totals$date, label = TRUE, abbr = TRUE)

US_totals$Season <- case_when(
  US_totals$month %in% c("Jan", "Feb", "Mar") ~ "Winter",
  US_totals$month %in% c("Apr", "May", "Jun") ~ "Spring",
  US_totals$month %in% c("Jul", "Aug", "Sep") ~ "Summer",
  US_totals$month %in% c("Oct", "Nov", "Dec") ~ "Fall"
)

US_totals$Year <- year(US_totals$date)

tail(US_totals)
```

***Timeframe of Covid-19 Cases and Deaths:***
The time frame graphs do not give a cumulative sum, but instead read the 
difference of new cases and deaths on one particular day compared to the day
before. The drops in the graph represent zero or small negative change. Note
the difference in scale between cases and deaths on the y-axis of the graphs
where difference in cases reach over 1,000,000 on one day compared to deaths where the 
maximum amount of difference of deaths between two days was over 4000. The red
line in the "New Deaths in US by Month" graph illustrates a slow growth of 
deaths per million of people.
```{r}
US_totals %>% ggplot() +
  geom_area(aes(x = date, y = new_cases, fill = Season)) +
  facet_wrap(~ Year, ncol = 1, scales = "free_x") +
  scale_x_date(date_labels = "%b") +
  scale_fill_manual(values = c(
    "Winter" = "#00EEEE",
    "Spring" = "#7FFF00",
    "Summer" = "#CD6600",
    "Fall" = "#9932CC"
  )) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "New Cases in US by Month",
       x = "Date",
       y = "New Cases") + 
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "#8B8B83"),
    panel.grid.minor = element_blank()
    )

US_totals %>% ggplot() +
  geom_area(aes(x = date, y = new_deaths, fill = Season)) +
  geom_line(aes(x = date, y = deaths_per_mill), color = "red", linewidth = 1) +
  facet_wrap(~ Year, ncol = 1, scales = "free_x") +
  scale_x_date(date_labels = "%b") +
  scale_fill_manual(values = c(
    "Winter" = "#00EEEE",
    "Spring" = "#7FFF00",
    "Summer" = "#CD6600",
    "Fall" = "#9932CC"
  )) +
  labs(title = "New Deaths in US by Month",
       x = "Date",
       y = "New Deaths") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.text.y = element_text(size = 6),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
        axis.title = element_text(size = 14, color = "#000000"),
        axis.text = element_text(size = 12, color = "#000000"),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "#8B8B83", color = NA)
        )
```

## Quantitative Boxplot Analysis:

```{r}
US_totals %>% ggplot() +
  geom_boxplot(aes(x = month, y = new_cases, fill = Season), linewidth = 0.9) +
  scale_fill_manual(values = c(
    "Winter" = "#00EEEE",
    "Spring" = "#7FFF00",
    "Summer" = "#CD6600",
    "Fall" = "#9932CC"
  )) + 
  scale_y_continuous(labels = scales::comma) +
  labs(title = "New Cases Quantitative Boxplots(2020-23)",
       x = "Date",
       y = "New Deaths") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    panel.grid.major = element_line(color = "#8B8B83"),
    panel.grid.minor = element_blank()
  )

summary(US_totals$new_cases)



US_totals %>% ggplot() +
  geom_boxplot(aes(x = month, y = new_deaths, fill = Season), linewidth = 0.9) +
  scale_fill_manual(values = c(
    "Winter" = "#00EEEE",
    "Spring" = "#7FFF00",
    "Summer" = "#CD6600",
    "Fall" = "#9932CC"
  )) + 
  labs(title = "New Deaths Quantitative Boxplots(2020-23)",
       x = "Date",
       y = "New Deaths") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
        axis.title = element_text(size = 14, color = "#000000"),
        axis.text = element_text(size = 12, color = "#000000"),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "#8B8B83", color = NA)
  )

summary(US_totals$new_deaths)
```

***New Cases and New Deaths Analysis:***

### New Cases
The day that had the greatest difference in new cases was +1,354,508 more cases 
on January 10, 2022 than January 9, 2022. Late fall 2021 - early winter 2022 
was a time of extreme COVID-19 spread. On average, the difference in new cases was about +90,896. 
This average is heavily influenced by outliers of days with
huge amounts of case differences especially during December, 2021 and Janurary, 2022.
Even when the vaccine was distributed in mid-2021, cases started to explode 
toward the end of 2021 as people in America were trying to return to their 
standard way of life.

### New Deaths
According to the new deaths boxplots, we can estimate the future death difference
between two days is +985 deaths. This number does not make sense at the beginning of the 
pandemic, but as the virus continued to spread, the amount of deaths would
steadily increase. The day with the greatest difference in deaths was January
12, 2021 with +4375 more people that died from COVID-19 compared to January 
11, 2021. January, Feburary, and December have the greatest variability of 
daily death differences as the colder weather can lead to lower sunlight and weakened immune
systems. There are many outliers in September and October due to a pattern 
similar to Influenza where as the weather starts to get colder and darker, 
cases and deaths steadily go up leading into winters. 

***Linear Model:***
If there is no difference in new cases presented on a day from the day before, we could estimate a death toll difference of about +648-649 people to die that same day. According to the linear estimates,
for every difference of +270 new cases adds +1 difference of deaths. The adjusted
R-squared value of 0.2531 indicates a weak, positive correlation between the daily difference
of cases and daily difference of deaths, but not enough to indicate a very 
strong correlation. This certainly would explain the extreme spread of the data
especially during seasonal changes. 

```{r}
mod2 <- lm(new_deaths ~ new_cases, data = US_totals)
mod2
summary(mod2)
```

***Project Bias:***
There are many instances of bias that we need to account for:

-Different states have different COVID protocols. We can not assume that the
method of measurements were the same throughout all states. Healthcare is 
different across all states as well.

-This dataset presents a huge emphasis on urban trends and areas. 
While rural areas are included, it is easy for these areas to be 
underrepresented.

-Considering that symptoms for some individuals registered as asymptomatic or
mild, it would be wrong to assume that there were no more cases outside of 
those reported in this study.


***Project Conclusion:***
January was the month with the greatest daily differences in COVID cases and 
the cold months of January, February, and December produce enormous variability
of daily death differences compared to warmer months. While the relationship 
in differences between COVID-19 cases and deaths was not correlated to a powerful
extent, the seasonal patterns align with extreme outliers as averages increase with
colder weather. While a difference in +200-300 cases does not necessarily 
predict how many people will die, it is still an important statistic for 
public health information to take into consideration when examining other 
lurking variables. While we certainly should not leave out the warmer months from early spring to earlier fall, we can safely say that winter/colder months, especially 
January 2021 and January 2022, had the greatest COVID-19 outbreaks. In the 
future, if more pandemics are to occur, we can at least try to focus our efforts 
toward stronger healthcare in colder months and give individuals incentives that
allow them to practice habits and activities to enhance their overall health.

```{r}
sessionInfo()
```